Biden Projected to Win the 2024 US Presidential Elections

### 2024-03-16 paper/paper.qmd
Discussion complete and typos fixed

---
title: "Biden Projected to Win the 2024 US Presidential Elections"
subtitle: "Analysis using data from American Community Survey 2022 and Ipsos Poll 2023 levereging Bayesian Logistic Regression with Post-Stratification"
author:
  - Monica Sainani
  - Kaavya Kalani
thanks: "Code and data supporting this analysis is available at: https://github.com/sainanim/Predictions-for-the-US-2024-Elections"
date: today
date-format: long
abstract: "We analyze American voting behavior, focusing on demographic, socioeconomic, and geographic factors. We use a Bayesian Logistic Regression model with Post-Stratification to predict the election outcome between Biden and Trump. We find a strong relationship between age, family income, education, state of residency, sex and likely voting outcome. Our central forecast is a Biden re-election. This research provides insights into the factors affecting political engagement in the U.S., potentially guiding future electoral strategies and policy discussions."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Our analysis led to us finding that overall, the younger generations, females and people of the midwest and southern states are highly supportive of the Democratic Party led by Biden. Our overall forecast is a Joe Biden re-election in the 2024 US Presidential Elections.

@@

Our survey data is from the Ipsos Poll [Roper #31120637]. It was collated from a web-based survey conducted on November 3 - 4, 2023 using KnowledgePanel. The survey is conducted by Ipsos and the sponsor is ABC News. This poll is based on a nationally representative probability sample of 949 adults age 18 or older with oversamples among Black and Hispanic respondents. The poll is a collection of demographic information and political preferences.

@@

- Education is the highest level of education attained by the panelist. They were given the option between: Less than high school, High school, Some college and Bachelor's degree or higher. We then classified these into three categories:  Less than high school, High school,  More than high school

@@

- Sex is the sex reported by the panelist and it is either male or female

For the variable we are predicting, we created a binary variable which is 1 if the panelist favors Biden and 0 if the panelist favors Trump. There was no question in the poll which asked this direct question so we used “Overall, do you have a favorable or unfavorable impression of...Donald Trump?”. We acknowledge that not favoring Donald Trump does not necessarily imply favoring Joe Biden, but since they are the two major potential candidates, we classify the panelists who chose “Unfavorable” as supporting Biden, i.e not supporting Trump.

@@

@fig-vote-biden shows the distribution of vote count in our survey data. We see that the survey panelists are heavily in favor of Joe Biden with a over twice of the panelists not favoring Trump. The bias can be a potential weakness of the dataset and we aim to uncover the factors influencing these political preferences.

@@

ggplot(state_proportions, aes(x = state, y = proportion, color = as.factor(vote_biden))) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_line(aes(group = vote_biden), position = position_dodge(width = 0.5)) +
  labs(x = "State", y = "Proportion", color = "Vote") +
  scale_color_manual(values = c("blue", "red"), labels = c("Vote for Trump", "Vote for Biden")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
@fig-variables shows the vote proportion distribution for our chosen variables of interests. We notice a noticeable difference in the proportions that favor Trump or Biden. Some striking trends we notice in our survey data are more male panelists supporting Trump, majority of Trump supporters being in the 45-59 age group and Trump support declining with increase in education levels. While most states seem to support Trump, we see that some states support Trump over Biden. This is a fresh change to see over all other factors showing a clear preference for Biden in all categories.

@@

# Plotting
ggplot(combined_data, aes(state)) +
  geom_point(aes(y = census_prop, color = "Census data"), size = 3) +
  geom_point(aes(y = sur_prop, color = "Survey data"), size = 3) +
  scale_color_manual(values = c("Census data" = "salmon", "Survey data" = "skyblue")) +
  labs(title = "State",
       x = "State",
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
@fig-proportions show a comparison of the survey and census data proportions. We notice that for all the variables of interest, the survey is not entirely representative of the population. For example, we see that females are underrepresented in the survey with the survey having about 49% females whereas the census shows 51% females. We also see how population with highest education attained being high school are under-represented whereas population with highest education attained being more than high school are over-representated. Difference in proportions like these can be seen in all the variables of interests and this is why post-stratifying is important to ensure precision and better representation of the population.

@@

The logistic regression model we will be using is:
\begin{equation}
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{state} + \beta_2 \times \text{sex} + \beta_3 \times \text{age} + \beta_4 \times \text{education} + \beta_5 \times \text{income} \label{eq:model1}
\end{equation}

@@

# Create the plot
ggplot(state_vote_for_biden, aes(y = biden_predict, x = reorder(state, -biden_predict), color = "MRP estimate")) +
  geom_point() +
  geom_errorbar(aes(ymin = biden_lower, ymax = biden_upper), width = 0) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    x = "Estimated proportion support for Biden",
    y = "State",
    color = "Source"
  ) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  coord_flip() +
  theme(legend.position = "bottom") +
  theme(axis.text.y = element_text(size = 4, margin = margin(0, 10, 0, 10))) # Adjust size of y-axis text

@@

The states that have points to the left of the black vertical line are the states voting for Trump,
and the states on the right are voting for Biden. We see that the majority of the votes for the democratic party
are found from the Midwest  and the Southern states while only Rhode Island, Oklahoma, Missouri and Nevada are inclined to vote for Trump.
Rhode Island is almost 75% more likely to vote for Trump, but in competition most of of Alaska, Hawaii, North Dakota, New Mexico and Kansas are inclined to vote for Biden.

@@

In @fig-support-age, we used the data to find the proportion of the voters categorized by age, who would vote for the Democratic party. We see that the younger generation are more likely to vote for Biden compared to the older. Americans under the age of 45 are 70% likely to vote for Trump. The support for Trump drops to around 60% in the 45-59 age group. While this is favorable towards Biden, it does indicate that more people between 45 to 59 years are likely to vote for Trump compared to other age groups.

@@

```
@fig-support-sex shows the proportion of Female voters that would vote for the Democratic Party as compared to the Male voters. We see a 10% difference in the proportion where women are 10% more likely to vote for Biden.

@@

@fig-support-general looks at the division of voters for each party as an overall. We see that even after applying post-stratification, the preference has only changed by a negligible amount with a large majority of the population likely to vote for Trump. This graph clearly shows the conclusion we can deduce from our research that it is clear that the population of the United States is more inclined towards the Democratic Party.

@@

## Political preference trends in different states 
In the graph, @fig-support-state, we notice that all states in the U.S. other than Rhode Island, Missouri, Oklahoma and Nevada are more inclined to vote for the Democratic Party. We see that as we go towards the Midwest and Southern Regions, the support for Biden increases as more than 50% of the state is likely to vote for them. This is similar to the findings in [@James_2016] that address the same prediction for 2024. Our model and the article predict that states that support Trump are from the middle and Midwest parts of the US, whereas coastal states and the Northeast tend to vote for Biden.

@@

The middle-aged generation i.e individuals of age group from 30-44 years old were inclined to voting for Biden as well. Middle-aged voters may align more closely with Biden's policy proposals on issues such as healthcare, economic recovery, climate change, and social justice [@Igielnik_2021] .

@@

## Contributions to Knowledge:

@@

Lastly, it highlights the high support of the Democratic Party led by Biden from the younger generations, females and people of the midwest and southern states.

@@

4. Education levels were standardized to facilitate analysis. Respondents' education information (ppeducat) was categorized into three groups: "High school", "More than high school" (including "Bachelors degree or higher" and "Some college"), and "Less than high school". This was done to match the education level separation available in the poststratification dataset. The final column was named education.

@@

@@

head(survey_data)
```

@@

head(poststrat_data)
```

>>>>

---
title: "Biden Projected to Win the 2024 US Presidential Elections"
subtitle: "Analysis using data from American Community Survey 2022 and Ipsos Poll 2023 leveraging Bayesian Logistic Regression with Post-Stratification"
author:
  - Monica Sainani
  - Kaavya Kalani
thanks: "Code and data supporting this analysis is available at: https://github.com/sainanim/Predictions-for-the-US-2024-Elections"
date: today
date-format: long
abstract: "We analyze American voting behavior, focusing on demographic, socioeconomic, and geographic factors. A Bayesian Logistic Regression model with Post-Stratification is used to predict the election outcome between Biden and Trump. We find a strong relationship between age, family income, education, state of residency, sex and likely voting outcome. Our central forecast is a Biden re-election. This research provides insights into the factors affecting political engagement in the U.S., potentially guiding future electoral strategies and policy discussions."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Our analysis led to us finding that overall, the younger generations, females, and people of the Midwest and southern states are highly supportive of the Democratic Party led by Biden. Our overall forecast is a Joe Biden re-election in the 2024 US Presidential Elections.

@@

Our survey data is from the Ipsos Poll [Roper #31120637]. It was collated from a web-based survey conducted on November 3 - 4, 2023 using KnowledgePanel. The survey is conducted by Ipsos and the sponsor is ABC News. This poll is based on a nationally representative probability sample of 949 adults aged 18 or older with oversamples among Black and Hispanic respondents. The poll is a collection of demographic information and political preferences.

@@

- Education is the highest level of education attained by the panelist. They were given the option between: Less than high school, High school, Some college and Bachelor's degree or higher. We then classified these into three categories:  Less than high school, High school,  More than high school.

@@

- Sex is the sex reported by the panelist and it is either male or female.

For the variable we are predicting, we created a binary variable which is 1 if the panelist favors Biden and 0 if the panelist favors Trump. There was no question in the poll which asked this direct question so we used “Overall, do you have a favorable or unfavorable impression of...Donald Trump?”. We acknowledge that not favoring Donald Trump does not necessarily imply favoring Joe Biden, but since they are the two major potential candidates, we classify the panelists who chose “Unfavorable” as supporting Biden, i.e. not supporting Trump.

@@

@fig-vote-biden shows the distribution of vote count in our survey data. We see that the survey panelists are heavily in favor of Joe Biden with over twice the panelists not favoring Trump. The bias can be a potential weakness of the dataset and we aim to uncover the factors influencing these political preferences.

@@

ggplot(state_proportions, aes(x = state, y = proportion, color = as.factor(vote_biden))) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_line(aes(group = vote_biden), position = position_dodge(width = 0.5)) +
  labs(x = "State", y = "Proportion", color = "Vote") +
  scale_color_manual(values = c("blue", "red"), labels = c("Vote for Trump", "Vote for Biden")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6, margin = margin(0, 10, 0, 10)))
```
@fig-variables shows the vote proportion distribution for our chosen variables of interests. We notice a noticeable difference in the proportions that favor Trump or Biden. Some striking trends we notice in our survey data are more male panelists supporting Trump, majority of Trump supporters being in the 45-59 age group and Trump support declining with increase in education levels. While most states seem to support Biden, we see that some states support Trump over Biden. This is a fresh change to see over all other factors showing a clear preference for Biden in all categories.

\newpage

@@

# Plotting
ggplot(combined_data, aes(state)) +
  geom_point(aes(y = census_prop, color = "Census data"), size = 3) +
  geom_point(aes(y = sur_prop, color = "Survey data"), size = 3) +
  scale_color_manual(values = c("Census data" = "salmon", "Survey data" = "skyblue")) +
  labs(title = "State",
       x = "State",
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6, margin = margin(0, 10, 0, 10)))

```
@fig-proportions show a comparison of the survey and census data proportions. We notice that for all the variables of interest, the survey is not entirely representative of the population. For example, we see that females are underrepresented in the survey with the survey having about 49% females whereas the census shows 51% females. We also see how population with highest education attained being high school are under-represented whereas population with highest education attained being more than high school are over-represented. Difference in proportions like these can be seen in all the variables of interests and this is why post-stratifying is important to ensure precision and better representation of the population.

\newpage

@@

The logistic regression model we will be using is:
\begin{equation}
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{state} + \beta_2 \times \text{sex} + \beta_3 \times \text{age} + \beta_4 \times \text{education} + \beta_5 \times \text{income}
\end{equation}

@@

# Create the plot
ggplot(state_vote_for_biden, aes(y = biden_predict, x = reorder(state, -biden_predict), color = "MRP estimate")) +
  geom_point() +
  geom_errorbar(aes(ymin = biden_lower, ymax = biden_upper), width = 0) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    x = "Estimated proportion support for Biden",
    y = "State",
    color = "Source"
  ) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  coord_flip() +
  theme(legend.position = "bottom") +
  theme(axis.text.y = element_text(size = 5, margin = margin(0, 10, 0, 10)))

@@

The states that have points to the left of the black vertical line are the states voting for Trump,
and the states on the right are voting for Biden. We see that the majority of the votes for the democratic party
are found from the Midwest and the southern states while only Rhode Island, Oklahoma, Missouri and Nevada are inclined to vote for Trump.
Rhode Island is almost 75% more likely to vote for Trump, but in competition, most of Alaska, Hawaii, North Dakota, New Mexico and Kansas are inclined to vote for Biden.

@@

In @fig-support-age, we see the proportion of the voters who would vote for the Democratic party categorized by age. We see that the younger generation are more likely to vote for Biden compared to the older. Americans under the age of 45 are 70% likely to vote for Trump. The support for Trump drops to around 60% in the 45-59 age group. While this is favorable towards Biden, it does indicate that more people between 45 to 59 years are likely to vote for Trump compared to other age groups.

@@

```
@fig-support-sex shows the proportion of female voters that would vote for the Democratic Party as compared to the male voters. We see a 10% difference in the proportion where women are 10% more likely to vote for Biden.

@@

@fig-support-general looks at the division of voters for each party as an overall. We see that even after applying post-stratification, the preference has only changed by a negligible amount with a large majority of the population likely to vote for Biden. This graph clearly shows the conclusion we can deduce from our research, the population of the United States is more inclined towards the Democratic Party.

@@

## Political preference trends in different states 
In the graph, @fig-support-state, we notice that all states in the U.S. other than Rhode Island, Missouri, Oklahoma and Nevada are more inclined to vote for the Democratic Party. We see that as we go towards the Midwest and southern regions, the support for Biden increases as more than 50% of the state is likely to vote for them. This is similar to the findings in [@James_2016] that address the same prediction for 2024. Our model and the article predict that states that support Trump are from the middle and Midwest parts of the US, whereas coastal states and the Northeast tend to vote for Biden.

@@

The middle-aged generation i.e. individuals of age group from 30-44 years old were inclined to voting for Biden as well. Middle-aged voters may align more closely with Biden's policy proposals on issues such as healthcare, economic recovery, climate change, and social justice [@Igielnik_2021].

@@

## Contributions to Knowledge

@@

Lastly, it highlights the high support of the Democratic Party led by Biden from the younger generations, females, and people of the Midwest and southern states.

## Weaknesses and Limitations

While our analysis provides valuable insights into voter behavior, it is important to acknowledge several limitations.

Firstly, our reliance on survey data from the Ipsos 2023 Poll introduces a sampling bias, as the sample does not fully represent the entire US population.

On analyzing the counts in the dataset, we see that some of the states have a count of 1 or 2 records in the dataset. For example, the District of Columbia and Wyoming, both have one record each in the dataset. While these states are not very populated so it is a given that the representation will be lesser, having only one representative leads to a very biased preference from that state. The political preference of the state will be considered as being what that one person’s preference is.

If we consider Alaska which has 2 votes, both the representatives from there are supportive of Biden leading to the biased conclusion that Alaska is highly supportive of the Democratic Party. This is a contradiction to the fact that Alaska is considered a state which is highly of the Republican Party [@Wikipedia_RedBlue].

Additionally, the survey data does not contain any participants from the state of Vermont and Montana. While these states are not very populated [@Wikipedia_2024], their opinions are still important to accurately analyze the preference splits throughout the country.

Consequently, these issues could lead to inaccuracies in our predictions and limit the generalizability of our findings to the broader population.

For our poststratification dataset from ACS, there was no dataset available from 2023. Our poststratification data relied on the survey that was conducted in 2022, and Americans' viewpoints on their choices for political parties may have changed in a year.

Additionally, the education variable from ACS had a vague description and only provided a certain level of education. The levels that were provided in the dataset were ranging from Grade 5 to 5+ years of college. This led to our education levels to be not very diverse but restricted to three major categories: Less than High School, High School and More than High School.

Lastly, while we focused on key demographic variables, other influential variables such as political ideology and campaign messaging were not fully accounted for in our model. Political parties use various data from the population such as media trends and ongoing cultural differences in the society. They also use micro-targeted political ads to voters on social media and online platforms [@Funakoshi_2020]. This means that our research has the inaccuracy of consideration of incomplete variable sets.

## Future Directions

Moving forward, further research is needed to address the limitations identified in our study and data to advance our understanding of electoral dynamics. Future studies could explore the integration of additional data sources, such as social media sentiment analysis and polling data, to enhance the accuracy of election predictions.

Moreover, there are new strategies opted by political parties in every election that are facilitated by new technologies. Exploring the role of media exposure, including news sources, social media platforms, and political messaging, in shaping citizens' attitudes, expectations, and priorities is something that should be considered in shaping the opinions of Americans.

@@

4. Education levels were standardized to facilitate analysis. Respondents' education information (ppeducat) was categorized into three groups: "High school", "More than high school" (including "Bachelor's degree or higher" and "Some college"), and "Less than high school". This was done to match the education level separation available in the poststratification dataset. The final column was named education.

@@

\newpage

@@

head(survey_data)  %>%
  kable(digits = 2)

@@

head(poststrat_data)  %>%
  kable(digits = 2)
```
